

\noindent 

\noindent 

\noindent 

\noindent 
\[1\] 


\noindent 

\noindent \textbf{Example 6.44}

\noindent  Example 6.44 Show that L ??$\mathrm{\{}$a${}^{n}$b${}^{n}$c${}^{n}$ where n \underbar{?}1$\mathrm{\}}$ is not context free.

\noindent Solution:

\noindent Step I: Assume that the language set L is a CFL. Let n be a natural number obtained by using the

\noindent pumping lemma.

\noindent Step II: Let z ??a${}^{n}$b${}^{n}$c${}^{n}$. So, {\textbar} z {\textbar} ??3n ??n. According to the pumping lemma for CFL, we can write

\noindent z ??uvwxy, where {\textbar} vx {\textbar} \underbar{?}1.

\noindent Step III: uvwxy ??a${}^{n}$b${}^{n}$c${}^{n}$ . As 1 \underbar{?}?{\textbar} vx {\textbar} \underbar{?}?n, ({\textbar} vwx {\textbar} \underbar{?}?n, so {\textbar} vx {\textbar} \underbar{?}?n) v or x cannot contain all the three symbols a, b, and c. So, v or x will be in any of the following forms.

\noindent 1. Contain only a and b, i.e., in the form a${}^{i}$b${}^{i}$.

\noindent 2. Or contain only b and c, i.e., in the form b${}^{i}$c${}^{i}$.

\noindent 3. Or contain only the repetition of any of the symbols among a, b, and c.

\noindent Let us take the value of k as 2. v${}^{2}$ or x${}^{2}$ will be in the form a${}^{i}$b${}^{i}$ a${}^{i}$b${}^{i}$ (as v is a string here aba is not equal

\noindent to a${}^{2}$ b or ba${}^{2}$) or b${}^{i}$c${}^{i}$ b${}^{i}$c${}^{i}$. So, uv${}^{2}$ wx${}^{2}$ y cannot be in the form a${}^{m}$b${}^{m}$c${}^{m}$. So, uv${}^{2}$ wx${}^{2}$ y $\mathrm{\notin }$?L.

\noindent If v or x contains repetition of any of the symbols among a, b, and c, then v or x will be any of the

\noindent form of a${}^{i\ }$,b${}^{i}$, or c${}^{i}$. Let us take the value of k as 0.uv${}^{0}$ wx${}^{0}$ y ??uwy. In the string, the number of occur-

\noindent of one of the other two symbols in uvy is less than n. So, uv${}^{2}$ wx${}^{2}$ y $\mathrm{\notin }$??L.

\noindent Example 6.45

\noindent  Example 6.45     Prove that the language L ??$\mathrm{\{}$a${}^{i2}$/i \underbar{?}?1 $\mathrm{\}}$ is not context free.

\noindent Solution:

\noindent Step I: Assume that the language set L is a CFL. Let n be a natural number obtained by using the

\noindent pumping lemma.

\noindent Step II: Let z ??a${}^{i2}$. So, {\textbar} z {\textbar} ??2${}^{i}$. Let 2${}^{i}$ ??n. According to the pumping lemma for CFL, we can write

\noindent z ??uvwxy, where {\textbar} vx {\textbar} \underbar{?}1 and {\textbar} vwx {\textbar} \underbar{?}?n.

\noindent Step III: The string z contains only `a', and so v and x will also be a string of only `a'. Let v ??ap and

\noindent x ??a${}^{q}$, where (p ??q) \underbar{?}?1. Since n \underbar{?}?0 and uvwxy ??2${}^{i}$, {\textbar} uv${}^{n}$wx${}^{n}$y {\textbar} ??{\textbar} uvwxy {\textbar} ??{\textbar} v${}^{n-1}$x${}^{n-1}$ {\textbar} ??2${}^{i}$ ??(p ??q)

\noindent (n ?1). As uv${}^{n}$wx${}^{n}$y $\mathrm{\in }$?L, {\textbar} uv${}^{n}$wx${}^{n}$y {\textbar} is also a power of 2, say 2${}^{j}$.

\noindent       (p ??q)(n ??1) ??2${}^{j}$ ??2${}^{i}$

\noindent $\mathrm{\ \ }\mathrm{\Rightarrow }$?(p ??q)(n ??1) ??2${}^{i}$ ??2${}^{j}$

\noindent $\mathrm{\ }\mathrm{\Rightarrow }$?(?(p ??q)2${}^{i}$${}^{\mathrm{\textrm{?}}}$${}^{1\ }$??2${}^{i}$ ??2${}^{j}$

\noindent  $\mathrm{\Rightarrow }$?(?2${}^{i}$ (2(p ??q) ??1) ??2${}^{j}$

\noindent (p ??q) may be even or odd, but 2(p ??q) is always even. However, 2(p ??q) ??1 is odd, which cannot be

\noindent a power of 2. Thus, L is not context free.

\noindent Example 6.46

\noindent Example 6.46 Prove that L = $\mathrm{\{}$0${}^{p}$/where p is prime$\mathrm{\}}$ is not a CFL.

\noindent Solution:

\noindent Step I: Suppose L ??L(G) is context free. Let n be a natural number obtained from the pumping lemma

\noindent for CFL.

\noindent 

\noindent 

\noindent 

\noindent 

\noindent \textbf{\textit{Step II: }}Let p be a number ??n, z ??0${}^{p}$ $\in $?L. By using the pumping lemma for CFL, we can write

\noindent z ??uvwxy, where {\textbar} vx {\textbar} $>$1 and {\textbar} vwx {\textbar} $\le \ $n.

\noindent \textbf{\textit{Step III: }}Let k ??0. From the pumping lemma for CFL, we can write uv${}^{0}$wx${}^{0}$y, i.e., uwy $\in \ $L. As uvy is

\noindent in the form 0${}^{p}$, where p is prime, {\textbar} uwy {\textbar} is also a prime number. Let us take it as q. Let {\textbar} vx {\textbar} ??r. Then,

\noindent {\textbar} uvqwxqy {\textbar} ??q ??qr ??q(1 ??r). This is not prime as it has factors q(1 ??r) including 1 and q(1 ??r). So,

\noindent uv${}^{q}$wx${}^{q}$y $\notin \ $L. This is a contradiction. Therefore, L is not context free.

\noindent \textbf{6.10 Ogden's Lemma for CFL}

\noindent If L is a context-free language, then there exists some positive integer p such that for any string w $\in $?L

\noindent with at least p positions marked, then w can be written as

\noindent w ??uvxyz

\noindent with strings u, v, x, y, and z, such that

\noindent 1. x has at least one marked position,

\noindent 2. either u and v both have marked positions or y and z both have marked positions,

\noindent 3. vxy has at most p marked positions, and

\noindent 4. uv${}^{i}$xy${}^{i}$z is in L for every i $\ge $?0.

\noindent Ogden's lemma can be used to show that certain languages are not context free, in cases where the

\noindent pumping lemma for CFLs is not suffi cient. An example is the language, $\mathrm{\{}$a${}^{i}$b${}^{j}$c${}^{k}$d${}^{l}$: i ??0 or j ??k ??l$\mathrm{\}}$. It is also useful to prove the inherent ambiguity of some languages.

\noindent \textbf{6.11 Decision Problems Related to CFG}

\noindent Decision problems are the problems which can be answered in a `yes' or `no'. There are a number of

\noindent decision problems related to CFG. The following are some of them.

\noindent     1. Whether a given CFG is empty or not. (Does it generate any terminal string?)

\noindent     2. Whether a given CFG is fi nite? (Does it generate up to a certain length word?)

\noindent     3. Whether a particular string w is generated by a given CFG. (Membership)

\noindent     4. Whether a given CFG is ambiguous.

\noindent     5. Whether two given CFG have a common word.

\noindent     6. Whether two different CFG generate the same language.

\noindent     7. Whether the complement of a given CFL is also a CFL.

\noindent   In the following section, we are discussing the fi rst three decision algorithms for CFG.

\noindent \textbf{6.11.1 Emptiness}

\noindent In a CFG $\mathrm{\{}$V${}_{N}$, $\mathit{\Sigma}$, P, S $\mathrm{\}}$, if S generates a string of terminal, then it is non-empty; otherwise, the cor-responding CFL is empty.

\noindent The following algorithm checks whether a CFG is empty or not:

\noindent While (there exist a production in the form NT${}_{i}$ $\to $?T/String of T)  

\noindent 

\noindent 

\noindent 

\noindent 

\noindent 

\noindent 

\noindent $\mathrm{\{}$

\noindent Replace each NT at the right hand side of each production by corresponding T or string of T.

\noindent $\mathrm{\}}$

\noindent if (S generates a string of terminal)

\noindent return non-empty.

\noindent else

\noindent empty.

\noindent \textbf{Example 6.47}

\noindent \textbf{Example 6.47 }Check whether the following CFG is empty or not.

\noindent S $\to $?AB/D

\noindent A $\to $?aBC

\noindent B $\to \ $bC

\noindent C $\to \ $d

\noindent D $\to $?CD

\noindent \textbf{\textit{Solution: }}C $\to $?d is in the form NT$\ \to \ $T. Replacing C $\to $?d in A $\to $?aBC, B $\to $?bC, and D$\ \to \ $CD, we get

\noindent S $\to $?AB/D

\noindent A $\to $aBd

\noindent B $\to $bd

\noindent D $\to $dD

\noindent B $\to $bd is in the form NT $\to $string of terminals. Replacing B $\to $bd in S $\to $AB and A $\to $aBd, we get

\noindent S $\to $Abd/D

\noindent A $\to $abdd

\noindent D $\to $dD

\noindent A $\to $abdd is in the form NT $\to $String of terminals. Replacing A $\to $abdd in S $\to $Abd, we get

\noindent S $\to $abddbd/D

\noindent D $\to $dD

\noindent As S generates a string of terminals, the CFG is non-empty.

\noindent \textbf{Example 6.48}

\noindent \textbf{Example 6.48 }Check whether the following CFG is empty or not.

\noindent S $\to $ABD

\noindent A $\to $BC

\noindent B $\to $aC

\noindent C $\to $b

\noindent D $\to $BD

\noindent \textbf{\textit{Solution: }}C $\to $b is in the form NT $\to $T. Replacing C $\to $b in A $\to $BC, B $\to $aC, we get

\noindent S $\to $ABD

\noindent A $\to $Bb

\noindent B $\to $ab

\noindent D $\to $BD

\noindent 

\noindent 

\noindent 

\noindent 

\noindent 

\noindent 

\noindent B $\to $ab is in the form NT $\to $String of T. Replacing B $\to $ab in S $\to $ABD, A $\to $Bb, D $\to $BD, we get

\noindent S $\to $AabD

\noindent A $\to $abb

\noindent D $\to $abD

\noindent A $\to $abb is in the form NT $\to $String of T. Replacing A $\to $abb in S $\to $AabD, we get

\noindent S $\to $abbabD

\noindent D $\to $abD

\noindent `While' part is complete, S does not produce any terminal string. Thus, the CFG is empty.

\noindent \textbf{6.11.2 Finiteness}

\noindent A language L generated from a given CFG is fi nite if there are no cycles in

\noindent the directed graph generated from the production rules of the given CFG. The

\noindent longest string generated by the grammar is determined by the derivation from

\noindent the start symbol.

\noindent \textit{(The number of vertices of the directed graph is the same as the number}

\noindent \textit{of non-terminals in the grammar. If there is a production rule S }$\to $\textit{AB, the}

\noindent \textit{directed graph is as shown in Fig. 6.15.)}

\noindent \textbf{\textit{6.11.2.1 Infiniteness}}

\noindent A language L generated from a given CFG is infi nite if there is at least one cycle in the directed graph

\noindent generated from the production rules of the given CFG.

\noindent The following examples describe this in detail.

\noindent \textbf{Example 6.49}

\noindent \textbf{Example 6.49 }Verify whether the languages generated by the following grammar are fi nite or not.

\noindent If fi nite, fi nd the longest string generated by the grammar. 

\noindent 

\noindent a) S $\to $AB  b) S $\to $AB

A $\to $BC   A $\to $B

B $\to $C   B $\to $SC/a

C $\to $a   C $\to $AB/b

\noindent 

\noindent \textbf{\textit{Soution:}}

\noindent a) The grammar is not in CNF. Removing the unit productions, the grammar becomes

\noindent S $\to $AB

\noindent A $\to $BC

\noindent B $\to $a

\noindent C $\to $a

\noindent Now the grammar is in CNF. The directed graph for the grammar is shown in Fig. 6.16.

